Open-Source machine learning BANTER acoustic classification of beaked whale echolocation pulses

### 2024-01-30 manuscript/CopyOfmanuscript.qmd
Revisions and Rebuttal
Revisions to manuscript + Rebuttal draft

The results for the four study areas varied from a low of 88.7% correct classification in the SAtlantic dataset to a high of 97.5% correct classification in the NAtlantic dataset. All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais' beaked whales (NAtlantic) had classification rates below 90%. Other studies on machine learning acoustic classification of beaked whales have yielded stronger classification results, although these studies included fewer beaked whale species classes and much larger sample sizes for each species class \[@li_model-based_2020-1, @lebien_species-level_2018, @cohen_identification_2022\]. While there is great room for improvement in our models, and we suspect that modest increases in sample sizes for some species will result in significant improvements to the model results, it is likely that the low sample sizes presented in these datasets would be insufficient to develop and test deep learning models.

@@

In addition to poor classification of Gervais' in the NAtlantic study area, the plot votes show that for one of the four events, only a few of the 100,000 trees 'voted' for Gervais', and the majority of the votes were for True's beaked whales. Analysts' notes indicated uncertainty in manual species classification of this event in this dataset. While small sample size can at times provide good classification scores, as we found for Sowerby's beaked whale in this SAtlantic dataset, they can be heavily influential if there are inaccuracies in the training data.

@@

The workflow presented here provides a highly automated approach to detection of acoustic events (PAMGuard), integration of environmental data (PAMpal), and acoustic event classification (BANTER). These methods significantly reduce manual analysis, provide more consistent classification results with fewer biases, and provide an estimate of classification error. The greatest improvement to classification results for beaked whales would likely result from improved sample sizes, and examination of individuals to accurately measure IPI. Consideration of additional detectors (e.g., matched filter detector, click train detector) or additional environmental variables may further improve classification results. Improved alignment of detection methods across studies (i.e., definition of events, hydrophone calibration) may allow for combining data within geographic regions to improve assessments within regions for all species, and across regions for species with global distributions, such as Cuvier's beaked whales. These highly automated methods may allow for analysis of data across large spatial and temporal scales to address large ecological and population level questions.

@@

# Figures and tables

```{r, natlantic}
#| label: fig-natlantic
#| fig-cap: BANTER classification results from the NAtlantic dataset including echolocation pulses and inter-pulse interval (EC_IPI). Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
#| fig-width: 10
#| fig-height: 10
#| fig-align: center
#| out-width: 100%
#| echo: false
confuse <- magick::image_read('manuscript_files/natlantic_ici_confuseMatrix.png')%>%
  image_border(color="#ffffff", geometry = "50x130")%>%
  image_annotate("a) Confusion Matrix", size=100, color = "black")
vote <- magick::image_read('manuscript_files/natlantic_ici_votes.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("d) Vote Plot", size=100, color = "black")
prox <- magick::image_read('manuscript_files/natlantic_ici_proximity.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("b) Proximity Plot", size=100, color = "black")
heat <- magick::image_read('manuscript_files/natlantic_ici_importance.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_scale("3300")%>%
  image_annotate("c) Importance Heat Map", size=100, color = "black")

@@

\newpage

```{r, satlantic}
#| label: fig-satlantic
#| fig-cap: BANTER classification results from the SAtlantic dataset (EC_IPI). Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
#| fig-width: 10
#| fig-height: 10
#| fig-align: center
#| out-width: 100%
#| echo: false
confuse <- magick::image_read('manuscript_files/satlantic_ici_confuseMatrix.png')%>%
  image_border(color="#ffffff", geometry = "50x130")%>%
  image_annotate("a) Confusion Matrix", size=100, color = "black")
vote <- magick::image_read('manuscript_files/satlantic_ici_votes.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("d) Vote Plot", size=100, color = "black")
prox <- magick::image_read('manuscript_files/satlantic_ici_proximity.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("b) Proximity Plot", size=100, color = "black")
heat <- magick::image_read('manuscript_files/satlantic_ici_importance.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_scale("3300")%>%
  image_annotate("c) Importance Heat Map", size=100, color = "black")

@@

\newpage

```{r, hawaii}
#| label: fig-hawaii
#| fig-cap: BANTER classification results from the Hawaii dataset (EC). Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
#| fig-width: 10
#| fig-height: 10
#| fig-align: center
#| out-width: 100%
#| echo: false
confuse <- magick::image_read('manuscript_files/hawaii_ec_confuseMatrix.png')%>%
  image_border(color="#ffffff", geometry = "50x130")%>%
  image_annotate("a) Confusion Matrix", size=100, color = "black")
vote <- magick::image_read('manuscript_files/hawaii_ec_votes.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("d) Vote Plot", size=100, color = "black")
prox <- magick::image_read('manuscript_files/hawaii_ec_proximity.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("b) Proximity Plot", size=100, color = "black")
heat <- magick::image_read('manuscript_files/hawaii_ec_importance.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_scale("3300")%>%
  image_annotate("c) Importance Heat Map", size=100, color = "black")

@@

\newpage

```{r, epacific-env}
#| label: fig-epacific
#| fig-cap: BANTER classification results from the EPacific dataset with environmental data (EC_IPI_ENV). Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
#| fig-width: 10
#| fig-height: 10
#| fig-align: center
#| out-width: 100%
#| echo: false
confuse <- magick::image_read('manuscript_files/epacific_env_confuseMatrix.png')%>%
  image_border(color="#ffffff", geometry = "50x130")%>%
  image_annotate("a) Confusion Matrix", size=100, color = "black")
vote <- magick::image_read('manuscript_files/epacific_env_votes.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("d) Vote Plot", size=100, color = "black")
prox <- magick::image_read('manuscript_files/epacific_env_proximity.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("b) Proximity Plot", size=100, color = "black")
heat <- magick::image_read('manuscript_files/epacific_env_importance.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_scale("3300")%>%
  image_annotate("c) Importance Heat Map", size=100, color = "black")

@@

\newpage

>>>>

Development of automated acoustic classifiers is critical for efficient analysis of large passive acoustic datasets; these classification routines require representative training data created by experienced acoustic analysts. While beaked whales species can be identified based on their stereotyped frequency-modulated echolocation clicks @baumann-pickering_species-specific_2013, not all acoustic events include sounds that can be readily attributed to a specific species. Moreover, there are a number of beaked whales where little or nothing is known about their acoustic repertoire and new species have been detected in recent years [@deangelis_description_2018, @barlow_unique_2022]. Therefore, development of acoustic classifiers for beaked whales should accommodate small and/or unbalanced sample sizes and have sufficient flexibility to allow for variable call characteristics.

The results for the four study areas varied from a low of 88.7% correct classification in the SAtlantic dataset to a high of 97.5% correct classification in the NAtlantic dataset. All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais' beaked whales (NAtlantic) had classification rates below 90%. Recent beaked whale classification studies using deep learning methods exhibit more robust results [@li_model-based_2020-1, @lebien_species-level_2018, @cohen_identification_2022, @ziegenhorn_discriminating_2022]; however, these models may not be appropriate for our dataset. Several studies included incomplete beaked whale species for the study area (@li_model-based_2020-1, @ziegenhorn_discriminating_2022 ) and all studies has sample large sample sizes (\>200) for each species class. The deep learning model developed by @cohen_identification_2022 included strong classification results with all beaked whale species found in our Atlantic models; however, it is unclear how this dataset trained on seafloor recorders would perform with towed array datasets. The BANTER models presented here only require a minimum of two samples per class and can therefore include all species within each region, making them useful to assess the presence of all beaked whales in the region. While deep learning methods may provide improved results once there are sufficient sample sizes for all relevant species within a study area, our results suggest that reasonable classification results can be obtained for all species using BANTER classification methods with as few as two samples per class.

@@

In addition to poor classification of Gervais' in the NAtlantic study area, the plot votes show that for one of the four events, only a few of the 100,000 trees 'voted' for Gervais', and the majority of the votes were for True's beaked whales. Analysts' notes indicated uncertainty in manual species classification of this event in this dataset. While small sample size can at times provide good classification scores, as we found for Sowerby's beaked whale in this SAtlantic dataset, they can be heavily influenced by inaccuracies in the training data.

@@

Until deep learning methods become more accessible and practical for applications involving rare species and low sample sizes, or their applicability across recording platforms can be guaranteed, machine learning approaches such as BANTER can serve as an accessible means of developing acoustic classifiers for complicated datasets. The workflow presented here provides a highly automated approach to detection of acoustic events (PAMGuard), integration of environmental data (PAMpal), and acoustic event classification (BANTER). These methods significantly reduce manual analysis, provide more consistent classification results with fewer biases, and provide an estimate of classification error. The greatest improvement to classification results for beaked whales would likely result from improved sample sizes, and examination of individuals to accurately measure IPI. Consideration of additional detectors (e.g., matched filter detector, click train detector) or additional environmental variables may further improve classification results. Improved alignment of detection methods across studies (i.e., definition of events, hydrophone calibration) may allow for combining data within geographic regions to improve assessments within regions for all species, and across regions for species with global distributions, such as Cuvier's beaked whales. These highly automated methods may allow for analysis of data across large spatial and temporal scales to address large ecological and population level questions.

@@

{{< pagebreak >}}
# Figures and tables

```{r, natlantic}
#| label: fig-natlantic
#| fig-cap: BANTER classification results from the NAtlantic dataset including echolocation pulses and inter-pulse interval (EC_IPI). Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
#| fig-width: 10
#| fig-height: 10
#| fig-align: center
#| out-width: 200%
#| echo: false
confuse <- magick::image_read('manuscript_files/natlantic_ici_confuseMatrix.png')%>%
  image_border(color="#ffffff", geometry = "80x208")%>%
  image_annotate("a) Confusion Matrix", size=120, color = "black")
vote <- magick::image_read('manuscript_files/natlantic_ici_votes.png')%>%
  image_border(color="#ffffff", geometry = "260x125")%>%
  image_annotate("d) Vote Plot", size=120, color = "black")
prox <- magick::image_read('manuscript_files/natlantic_ici_proximity.png')%>%
  image_border(color="#ffffff", geometry = "260x125")%>%
  image_annotate("b) Proximity Plot", size=120, color = "black")
heat <- magick::image_read('manuscript_files/natlantic_ici_importance.png')%>%
  image_border(color="#ffffff", geometry = "110x260")%>%
  image_scale("3300")%>%
  image_annotate("c) Importance Heat Map", size=120, color = "black")

@@

{{< pagebreak >}}

```{r, satlantic}
#| label: fig-satlantic
#| fig-cap: BANTER classification results from the SAtlantic dataset (EC_IPI). Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
#| fig-width: 10
#| fig-height: 10
#| fig-align: center
#| out-width: 200%
#| echo: false
confuse <- magick::image_read('manuscript_files/satlantic_ici_confuseMatrix.png')%>%
  image_border(color="#ffffff", geometry = "80x208")%>%
  image_annotate("a) Confusion Matrix", size=120, color = "black")
vote <- magick::image_read('manuscript_files/satlantic_ici_votes.png')%>%
  image_border(color="#ffffff", geometry = "260x125")%>%
  image_annotate("d) Vote Plot", size=120, color = "black")
prox <- magick::image_read('manuscript_files/satlantic_ici_proximity.png')%>%
  image_border(color="#ffffff", geometry = "260x125")%>%
  image_annotate("b) Proximity Plot", size=120, color = "black")
heat <- magick::image_read('manuscript_files/satlantic_ici_importance.png')%>%
  #image_border(color="#ffffff", geometry = "540x52")%>%
  image_border(color="#ffffff", geometry = "110x260")%>%
  image_scale("3300")%>%
  image_annotate("c) Importance Heat Map", size=120, color = "black")

@@

{{< pagebreak >}}

```{r, hawaii}
#| label: fig-hawaii
#| fig-cap: BANTER classification results from the Hawaii dataset (EC). Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
#| fig-width: 10
#| fig-height: 10
#| fig-align: center
#| out-width: 200%
#| echo: false
confuse <- magick::image_read('manuscript_files/hawaii_ec_confuseMatrix.png')%>%
  image_border(color="#ffffff", geometry = "80x208")%>%
  image_annotate("a) Confusion Matrix", size=120, color = "black")
vote <- magick::image_read('manuscript_files/hawaii_ec_votes.png')%>%
  image_border(color="#ffffff", geometry = "260x125")%>%
  image_annotate("d) Vote Plot", size=120, color = "black")
prox <- magick::image_read('manuscript_files/hawaii_ec_proximity.png')%>%
  image_border(color="#ffffff", geometry = "260x125")%>%
  image_annotate("b) Proximity Plot", size=120, color = "black")
heat <- magick::image_read('manuscript_files/hawaii_ec_importance.png')%>%
  image_border(color="#ffffff", geometry = "110x260")%>%
  image_scale("3300")%>%
  image_annotate("c) Importance Heat Map", size=120, color = "black")

@@

{{< pagebreak >}}

```{r, epacific-env}
#| label: fig-epacific
#| fig-cap: BANTER classification results from the EPacific dataset with environmental data (EC_IPI_ENV). Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
#| fig-width: 10
#| fig-height: 10
#| fig-align: center
#| out-width: 200%
#| echo: false
confuse <- magick::image_read('manuscript_files/epacific_env_confuseMatrix.png')%>%
  image_border(color="#ffffff", geometry = "80x208")%>%
  image_annotate("a) Confusion Matrix", size=120, color = "black")
vote <- magick::image_read('manuscript_files/epacific_env_votes.png')%>%
  image_border(color="#ffffff", geometry = "260x125")%>%
  image_annotate("d) Vote Plot", size=120, color = "black")
prox <- magick::image_read('manuscript_files/epacific_env_proximity.png')%>%
  image_border(color="#ffffff", geometry = "260x125")%>%
  image_annotate("b) Proximity Plot", size=120, color = "black")
heat <- magick::image_read('manuscript_files/epacific_env_importance.png')%>%
  image_border(color="#ffffff", geometry = "110x260")%>%
  image_scale("3300")%>%
  image_annotate("c) Importance Heat Map", size=120, color = "black")

@@

{{< pagebreak >}}
